\documentclass[a4paper]{article}
\usepackage{amsmath}

\parindent=0pt
\setcounter{secnumdepth}{2}

\newcommand{\N}[1]{\overline{#1\vphantom{l}}}
\renewcommand{\L}{{\leftarrow}}
\newcommand{\R}{{\rightarrow}}
\newcommand{\U}{{\uparrow}}
\newcommand{\D}{{\downarrow}}

\begin{document}

\section*{Solving a Yajilin problem with a SAT solver}
\begin{flushright}
2018.9.12 y-takata
\end{flushright}

Yajilin's rule and how to play it: \\
https://www.nikoli.co.jp/en/puzzles/yajilin.html

\section{Preliminaries}
Let $W$ and $H$ be the width and height of the game board, respectively.
``Cell $(i,j)$'' denotes the cell at the $i$th column and the $j$th row
($1\le i\le W$, $1\le j\le H$).
%
Let $X=\{,\R,\U,\D\}$.
Each element of $X$ is a function from a cell to a cell:
e.g., $\R(i,j)=(i+1,j)$.

\section{SAT encoding}
\subsection{Propositional variables}
\begin{itemize}
\item $b_{i,j}$: Cell $(i,j)$ is black.
\item $d_{i,j}$: Cell $(i,j)$ contains a number.
\item $l_{i,j,a}$:
  Cells $(i,j)$ and $a(i,j)$ are linked with a line ($a\in X$).
\item $c_{i,j,x,y,m}$:
  From cell $(i,j)$, you can go to cell $(x,y)$ along the lines
  within $m$ steps.
\item $c'_{i,j,x,y,m,a}$:
  Auxiliary variables for representing $c_{i,j,x,y,m}\land l_{x,y,a}$.
\item $n_{i,j,a,m}$:
  There are at least $m$ black cells in the direction $a$ of
  cell $(i,j)$ (excluding the cell $(i,j)$).
\end{itemize}

\subsection{Puzzle rules}
\subsubsection{Basic rules}
\begin{itemize}
\item Cells with numbers never be black. \
  $\N{d_{i,j}}\lor\N{b_{i,j}}$.

\item Black cells do not contain a line. \
  $\N{b_{i,j}}\lor\N{l_{i,j,a}}$.

\item Cells with numbers does not contain a line. \
  $\N{d_{i,j}}\lor\N{l_{i,j,a}}$.

\item
  Two black cells never touch horizontally or vertically.
  $
    (\N{b_{i,j}}\lor\N{b_{i+1,j}})
   \ {}\land{}\
    (\N{b_{i,j}}\lor\N{b_{i,j+1}}).
  $

\item
  If cell $(i,j)$ contains a number, then $d_{i,j}$.
\item
  If cell $(i,j)$ does not contain a number, then $\N{d_{i,j}}$.
\end{itemize}

\subsubsection{Lines}

\begin{itemize}
\item The link between cells are symmetric.
  \par\qquad
  $\begin{alignedat}[t]{4}
   & (\N{l_{i,j,\R}} &&\lor   l_{i+1,j,\L})  &&\land
     (   l_{i,j,\R}  &&\lor\N{l_{i+1,j,\L}}) \\ {}\land{}
   & (\N{l_{i,j,\D}} &&\lor   l_{i,j+1,\U})  &&\land
     (   l_{i,j,\D}  &&\lor\N{l_{i,j+1,\U}}).
  \end{alignedat}$

\item No line goes outside of the board. \
  $\begin{alignedat}[t]{3}
   \text{(horizontally)}\quad &
   \N{l_{1,j,\L}} &&\land \N{l_{W,j,\R}} \\
   \text{(vertically)}\quad &
   \N{l_{i,1,\U}} &&\land \N{l_{i,H,\D}}
   \end{alignedat}$

\item
  Every cell has at most two linked cells.  \
  $\begin{alignedat}[t]{3}
   & (\N{l_{i,j,\L}}&&\lor\N{l_{i,j,\R}}&&\lor\N{l_{i,j,\U}}) \\ {}\land{}
   & (\N{l_{i,j,\L}}&&\lor\N{l_{i,j,\R}}&&\lor\N{l_{i,j,\D}}) \\ {}\land{}
   & (\N{l_{i,j,\L}}&&\lor\N{l_{i,j,\U}}&&\lor\N{l_{i,j,\D}}) \\ {}\land{}
   & (\N{l_{i,j,\R}}&&\lor\N{l_{i,j,\U}}&&\lor\N{l_{i,j,\D}}).
  \end{alignedat}$

\item
  No cell is linked with exactly one cell. \
  $\begin{alignedat}[t]{1}
   &(l_{i,j,\L}\lor l_{i,j,\R}\lor l_{i,j,\U}\lor\N{l_{i,j,\D}})\\{}\land{}
   &(l_{i,j,\L}\lor l_{i,j,\R}\lor\N{l_{i,j,\U}}\lor l_{i,j,\D})\\{}\land{}
   &(l_{i,j,\L}\lor\N{l_{i,j,\R}}\lor l_{i,j,\U}\lor l_{i,j,\D})\\{}\land{}
   &(\N{l_{i,j,\L}}\lor l_{i,j,\R}\lor l_{i,j,\U}\lor l_{i,j,\D}).
  \end{alignedat}$
\end{itemize}

\subsubsection{The lines form a single loop}

\begin{itemize}
\item
  Every cell is reachable from itself within zero step, and
  no cell is reachable from another cell within zero step. \
  $c_{i,j,i,j,0}\land \bigwedge_{(x,y)\ne(i,j)} \N{c_{i,j,x,y,0}}$.

\item
  $c'_{i,j,x,y,m,a}$ is equivalent to $c_{i,j,x,y,m}\land l_{x,y,a}$.
  \par\qquad
  $
%  &(\N{c_{i,j,x,y,m}}\lor\N{l_{x,y,a}}\lor{c'_{i,j,x,y,m,a}})\\{}\land{}
   (\N{c'_{i,j,x,y,m,a}}\lor{c_{i,j,x,y,m}})\ {}\land{}\
   (\N{c'_{i,j,x,y,m,a}}\lor{l_{x,y,a}}).
   $

\item
  A cell is reachable from cell $(i,j)$ within $m+1$ steps
  only if it is reachable within $m$ steps or
  it is linked to a cell that is reachable within $m$ steps.
   \par\qquad
     $\begin{alignedat}[t]{2}
      \N{c_{i,j,x,y,m+1}}\lor
      c_{i,j,x,y,m} &\lor
      c'_{i,j,\R(x,y),m,\L} &&\lor
      c'_{i,j,\D(x,y),m,\U} \\ &\lor
      c'_{i,j,\L(x,y),m,\R} &&\lor
      c'_{i,j,\U(x,y),m,\D}.
      \end{alignedat}$

\item
  Every two cells that are not black and do not have numbers
  must be reachable from each other.
  (The number of steps between those two cells
  is at most the half of the length of the loop.)
  \begin{equation}
   d_{i,j}\lor b_{i,j}\lor
   d_{x,y}\lor b_{x,y}\lor
   c_{i,j,x,y,\lfloor WH/2\rfloor}.
   \label{eq:reachability}
  \end{equation}
\end{itemize}

\subsubsection{The number of black cells}

\begin{itemize}
\item
  No black cell in the left of cell $(1,j)$. \
$n_{1,j,\L,0}
\land \N{n_{1,j,\L,1}} \land\ldots\land \N{n_{1,j,\L,W}}$.

\item
  The number of black cells in the left of cell $(i+1,j)$.
  \par\qquad
  $\begin{aligned}[t]
   &(\N{n_{i,j,\L,m}}\lor n_{i+1,j,\L,m})\\{}\land{}
   &(\N{n_{i,j,\L,m-1}}\lor\N{b_{i,j}}\lor n_{i+1,j,\L,m})\\{}\land{}
   &(\N{n_{i+1,j,\L,m}}\lor n_{i,j,\L,m}\lor n_{i,j,\L,m-1})\\
   {}\land{}
   &(\N{n_{i+1,j,\L,m}}\lor n_{i,j,\L,m}\lor b_{i,j}).
   \end{aligned}$

\item Similar clauses for the other directions.

\item
  Cell $(i,j)$ has an arrow $a$ and a number $k$. \
  $n_{i,j,a,k}\land\N{n_{i,j,a,k+1}}$.
\end{itemize}

\section{Reducing the number of clauses}

(In the following, ``an empty cell'' means a cell that is not black
and does not contain a number.)

\newcommand{\id}{\mathit{id}}
\advance\parskip by2pt
\parindent=1em

\subsection{The ``pivot'' cell}

It is sufficient to check each empty cell
is reachable from a single ``pivot'' cell, instead of
checking every two empty cells are reachable
from each other.
In other words,
we can fix a pair $i,j$ for
the variables $c_{i,j,x,y,m}$ and $c'_{i,j,x,y,m,a}$ to
some single coordinate.
By doing this, the number of variables and clauses
is reduced to $O((WH)^2)$.

Note that the pivot cell must be a cell without a number
that \emph{never} be black
(e.g., a cell at which a cell with $0$ is pointing,
a cell next to a cell that must be black, etc.).
We have to find any of such cells before the SAT encoding,
and then we generate the clauses only for that pivot cell.
(If we cannot find such a pivot cell, then
we cannot apply this optimization.
However, most Yajilin problems contain cells with $0$,
and so this optimization may be applicable in many cases.)


\subsection{Order of the cells}

The method in this section is meaningful only when
the above \emph{pivot-cell} method is not applicable.

Give a serial number from $1$ to $WH$ to each cell.
Let $\id(i,j)$ denote the serial number of cell $(i,j)$.
We define the variables $c_{i,j,x,y,m}$ and $c'_{i,j,x,y,m,a}$
only when $\id(i,j)\le\id(x,y)$.
By doing this, we can reduce the number of these variables and
related clauses by half.
%
However, the meaning of $c_{i,j,x,y,m}$ changes into:
``Cell $(x,y)$ is reachable from cell $(i,j)$ within $m$ steps
by going through only cells whose serial numbers are greater than or equal to
$\id(i,j)$.''
So $c_{i,j,x,y,\lfloor WH/2\rfloor}$ is not necessarily true when
cell $(x,y)$ is reachable from cell $(i,j)$ along the lines.
Hence, we introduce new propositional variables $p_k$ and
modify the clause~(\ref{eq:reachability}) as follows:

\begin{itemize}
\item $p_k$:
  There exists an empty cell whose serial number is less than or equal to $k$.
\item
  Every empty cell is reachable
  from the empty cell with the least serial number.
  \par\qquad
  $\N{p_{\id(i,j)}}\lor{p_{\id(i,j)-1}}\lor
   d_{x,y}\lor b_{x,y}\lor
   c_{i,j,x,y,\lfloor WH/2\rfloor}$.
\end{itemize}
%
The truth of $p_k$ is defined as follows:
\begin{itemize}
\item
  No empty cell whose serial number is zero. \
  $\N{p_0}$.
\item
  There exists an empty cell whose serial number is less than or equal to
  $k$ if and only if there exists an empty cell whose serial number
  is less than $k$ or the $k$th cell is empty.
  \par\qquad
  $\begin{aligned}[t]
   &(\N{p_{\id(i,j)-1}}\lor p_{\id(i,j)}) \\ {}\land{}
   &(b_{i,j}\lor d_{i,j}\lor p_{\id(i,j)}) \\ {}\land{}
   &(\N{p_{\id(i,j)}}\lor p_{\id(i,j)-1}\lor \N{b_{i,j}}) \\ {}\land{}
   &(\N{p_{\id(i,j)}}\lor p_{\id(i,j)-1}\lor \N{d_{i,j}}).
   \end{aligned}$
\end{itemize}

\subsection{Parity of the cells}

We call cell $(i,j)$ ``even'' if $i+j$ is even and
call it ``odd'' otherwise.
The number of steps between cells $(i,j)$ and $(x,y)$
along to the lines must be even if and only if
the parity of the two cells are equal.
So we define the variables $c_{i,j,x,y,m}$ and $c'_{i,j,x,y,m,a}$
only for even $m$ when the parity of $(i,j)$ equals that of $(x,y)$
and only for odd $m$ otherwise.
By doing this, we can reduce the number of these variables and
related clauses by half.

We can apply this optimization
in parallel with the \emph{pivot-cell} method
and the \emph{order-of-cells} method.

\end{document}
